---
title: "Species selection"
author: "Mike O'Brien"
date: '2021-02-16'
output:
  pdf_document:
    toc: true
    toc_depth: 4
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  dev = "png"
)
```

## Goals of this document

-   Find key demersal fish species caught in the NEFSC bottom trawl
    surveys within the Mid-Atlantic Bight and Southern New England
    strata

-   Visualize the distribution of the above species

-   Visualize the distribution of species noted in @rose2018 concept
    paper for Lenfest

    -   summer flounder
    -   black sea bass
    -   scup

## Key species within the MAB/SNE

The NEFSC bottom trawl survey is divvied up into strata. I'm going to
select those strata in what I think of as the "Mid-Atlantic Bight":
strata that are designated as from the mid-Atlantic or southern New
England:

```{r echo=FALSE, message=FALSE}
library(data.table)

strata_info <- fread('data/svdbs_supporttables/svdbs_svmstrata.csv')

library(sf); library(dplyr)
strata <- read_sf('data/mapping/strata/strata.shp') %>% 
  left_join(strata_info, by = c('STRATA' = 'stratum'))


coast <- read_sf('data derived/mapping/coast_crop.shp') %>% 
  st_transform(st_crs(strata))


library(ggplot2)

ggplot() + 
  geom_sf(data = coast) +
  geom_sf(data = strata,
          aes(color = case_when(grepl('MAB', stratum_name) ~ 'MAB',
                                grepl('SNE', stratum_name) ~ 'SNE',
                                T ~ 'Other'))) +
  coord_sf(xlim = c(-80, -65), ylim = c(32, 45)) +
  labs(color = NULL) +
  theme_bw() +
  theme(legend.position = c(0.8, 0.3))
```

In the MAB/SNE, the top 10 species according to incidence, abundance, or
biomass are:

```{r echo=FALSE}
survdat <- fread('data derived/survdat_names_sed.csv')


n_mab_sne <- survdat[stratum %in% strata_info[grepl('MAB|SNE', stratum_name)]$stratum]
n_mab_sne <- n_mab_sne[, lapply(.SD, sum, na.rm  = T),
                       .SDcols = c('abundance', 'biomass'),
                       by = 'comname']
n_mab_sne <- n_mab_sne[survdat[, .N, by = 'comname'], on = 'comname']

n_mab_sne[, ':='(rank_incid = frank(-N),
                 rank_abun = frank(-abundance),
                 rank_bio = frank(-biomass))]
n_mab_sne[, mn_rank := (rank_incid + rank_abun + rank_bio) / 3]

n_mab_sne[order(mn_rank), .(comname, rank_incid, rank_abun, rank_bio)][
  rank_incid < 10 | rank_abun < 10 | rank_bio < 10]
```

Of those species, the following are demersal according to
[FishBase](http://fishbase.org/):

-   little skate
-   silver hake
-   winter skate
-   red hake
-   scup
-   haddock
-   weakfish
-   Atlantic croaker
-   smooth dogfish
-   American lobster (though not a fish)
-   American plaice

If we limit the search to "MAB"-designated strata (i.e., those south of
Hudson Canyon and north of Cape Hatteras), our top-10 species are:

```{r echo=FALSE}
n_mab <- survdat[stratum %in% strata_info[grepl('MAB', stratum_name)]$stratum]
n_mab <- n_mab[, lapply(.SD, sum, na.rm  = T),
               .SDcols = c('abundance', 'biomass'),
               by = 'comname']
n_mab <- n_mab[survdat[, .N, by = 'comname'], on = 'comname']

n_mab[, ':='(rank_incid = frank(-N),
             rank_abun = frank(-abundance),
             rank_bio = frank(-biomass))]
n_mab[, mn_rank := (rank_incid + rank_abun + rank_bio) / 3]

n_mab[order(mn_rank), .(comname, rank_incid, rank_abun, rank_bio)][
  rank_incid < 10 | rank_abun < 10 | rank_bio < 10]
```

Of those species, the following are demersal:

-   little skate
-   spotted hake
-   northern searobin
-   silver hake
-   scup
-   weakfish
-   smooth dogfish
-   Atlantic croaker
-   red hake
-   winter skate
-   bluntnose stingray
-   American lobster (though, again, not a fish)
-   haddock
-   American plaice

## Spatial distribution

```{r echo=FALSE}
survdat_sp <- st_as_sf(survdat[comname %in%
                                 c(n_mab[rank_incid < 10 | rank_abun < 10 | rank_bio < 10, comname],
                                  c('summer flounder', 'black sea bass', 'scup'))&
                                 abundance > 0],
                       coords = c('lon', 'lat'),
                       crs = 4326) %>% 
  st_transform(st_crs(strata))

all_plots <- function(species){
  plot_func <- function(species, season, type){
    ggplot() +
      geom_sf(data = coast) +
      geom_sf(data = survdat_sp[survdat_sp$comname == species &
                                  survdat_sp$year %in% seq(1963, 2019, by = 5) &
                                  survdat_sp$season == toupper(season),],
              aes_(color = as.name(type))) +
      scale_color_viridis_c(name = paste0(toupper(substr(type, 1, 1)),
                                 tolower(substr(type, 2, nchar(type)))), 
                            trans = 'log10') +
      coord_sf(xlim = c(-76.5, -67), ylim = c(35, 43)) +
      labs(title = paste(paste0(toupper(substr(species, 1, 1)),
                                 tolower(substr(species, 2, nchar(species)))),
                          paste0(type, ':'),
                          tolower(season),
                          'trawls', sep = ' ')) +
      facet_wrap(~ year)
  }
  
  print(plot_func(species, 'spring', 'abundance'))
  
  print(plot_func(species, 'fall', 'abundance'))

  print(plot_func(species, 'spring', 'biomass'))

  print(plot_func(species, 'fall', 'biomass'))
  
}
```

### Species noted in @rose2018 Lenfest concept paper

#### Summer flounder

```{r echo=FALSE, warning=FALSE}
all_plots('summer flounder')
```

#### Black sea bass

```{r echo=FALSE, warning=FALSE}
all_plots('black sea bass')
```

#### Scup

```{r echo=FALSE, warning=FALSE}
all_plots('scup')
```

### Identified species

#### Little skate

```{r echo=FALSE}
all_plots('little skate')
```

#### Spotted hake

```{r echo=FALSE}
all_plots('spotted hake')
```

#### Northern searobin

```{r echo=FALSE}
all_plots('northern searobin')
```

#### Silver hake

```{r echo=FALSE}
all_plots('silver hake')
```

#### Weakfish

```{r echo=FALSE}
all_plots('weakfish')
```

#### Smooth dogfish

```{r echo=FALSE}
all_plots('smooth dogfish')
```

#### Atl. croaker

```{r echo=FALSE}
all_plots('atlantic croaker')
```

#### Red hake

```{r echo=FALSE}
all_plots('red hake')
```

#### Winter skate

```{r echo=FALSE}
all_plots('winter skate')
```

#### Bluntnose stingray

```{r echo=FALSE}
all_plots('bluntnose stingray')
```

#### American lobster

```{r echo=FALSE}
all_plots('american lobster')
```

#### Haddock

```{r echo=FALSE}
all_plots('haddock')
```

#### American plaice

```{r echo=FALSE}
all_plots('american plaice')
```

## References
