---
title: "First blush at species distribution modeling"
author: "Mike O'Brien"
date: 'Week of 2021-02-15'
output:
  pdf_document:
    toc: true
    toc_depth: 4
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  dev = "png", dpi = 144
)
```

## Goals of this document

-   Begin species distribution modeling (SDM)
-   Utilize Atlantic croaker as the model species
-   Start with GAMMs (what I'm most familiar with)
-   Unfortunately, metrics used in the [`sdm`](https://cran.r-project.org/web/packages/sdm/index.html) [@naimi2016], [`dismo`](https://cran.r-project.org/web/packages/dismo/index.html), and [`biomod2`](https://github.com/biomodhub/biomod2) [@thuiller2009] R packages are not working for me at the moment. Work in the future will use these.

## Atlantic croaker distribution

Here are all the trawls that dragged up a croaker through the years:

```{r}
library(ggplot2); library(sf); library(data.table)

coast <- read_sf('data derived/mapping/coast_crop.shp') %>% 
  st_transform(crs = '+proj=aea +lat_1=33 +lat_2=45 +lon_0=-74')

all_data <- fread('data derived/survdat_names_sed.csv')

station_key <- unique(all_data,
                      by = c('cruise6', 'station', 'stratum', 'tow',
                             'year', 'season'))
station_key[, 18:25 := NULL]

croaker <- all_data[comname == 'atlantic croaker']
croaker <- croaker[station_key, on = names(station_key)]
croaker <- unique(croaker, by = c('cruise6', 'station', 'stratum', 'tow',
                                  'year', 'season'))


croaker[, abundance := fifelse(is.na(abundance), 0, abundance)]

croaker <- st_as_sf(croaker,
                    coords = c('lon', 'lat'),
                    remove = F,
                    crs = 4326) %>% 
  st_transform(crs = '+proj=aea +lat_1=33 +lat_2=45 +lon_0=-74')


ggplot() +
  geom_sf(data = coast, color = 'lightgray') +
  geom_sf(data = croaker[croaker$abundance > 0,], alpha = 0.5) +
  theme_minimal()
```

And here they are split out according to time (and unfortunately over-plotted):

```{r}
setDT(croaker)[, cut_7y := cut(year, seq(1963, 2019, 7))]
croaker[, color.index := as.factor((year - 1963) %% 7)]

ggplot() + 
  geom_sf(data = coast, color = 'lightgray') +
  geom_sf(data = croaker[abundance > 0],
          aes(geometry = geometry, color = color.index),
          alpha = 0.3) +
  labs(color = 'Year in interval') +
  facet_wrap(~ cut_7y, ncol = 4) +
  theme_minimal()

```

## Restricting the model bounds

I'm going to restrict the data we're using to the MAB/SNE strata (pink), though this may change in the future if we choose to not explicitly include space in the models.

```{r}
strata_info <- fread('data/svdbs_supporttables/svdbs_svmstrata.csv')

strata <- read_sf('data/mapping/strata/strata.shp')
setDT(strata)

strata <- strata[strata_info, on = c('STRATA' = 'stratum')]
strata <- strata[grepl('MAB|SNE', stratum_name)]
strata[, geometry :=
         st_transform(geometry, crs = '+proj=aea +lat_1=33 +lat_2=45 +lon_0=-74')]

strata <- strata[, .(geometry = st_union(geometry))]

ggplot() +
  geom_sf(data = coast, color = 'lightgray') +
    geom_sf(data = croaker[abundance > 0], aes(geometry = geometry)) +
  geom_sf(data = strata, aes(geometry = geometry),
          fill = 'pink', alpha = 0.5) +
  theme_minimal()
```

This inclusion rule causes the sites below to be dropped or selected. Note that we are dropping some sited that are located within Delaware Bay, which tends to be an Atlantic croaker hot spot.

```{r}
croaker[, selected := as.vector(st_intersects(geometry,
                                                 strata$geometry,
                                                 sparse = F))]
croaker[, selected := fifelse(selected, 'Selected', 'Dropped')]

ggplot() +
  geom_sf(data = coast, color = 'lightgray') +
    geom_sf(data = strata, aes(geometry = geometry),
          fill = 'pink', alpha = 0.5) +
    geom_sf(data = croaker[abundance > 0], aes(geometry = geometry)) +

  facet_wrap(~ selected) +
  theme_minimal()
```

Below is the distribution of croaker within the MAB/SNE strata, summed within 7 year blocks. This is the data I'll be using for the model (though the actual observations, not summed within blocks).

```{r}
yrly7 <- croaker[selected == 'Selected', 
                 .(abundance = sum(abundance, na.rm = T),
                   x = mean(st_coordinates(geometry)[,1]),
                   y = mean(st_coordinates(geometry)[,2])),
                 by = c('station', 'stratum', 'tow', 'cut_7y')]

ggplot() +
  geom_sf(data = coast, color = 'lightgray') +
  geom_point(data = yrly7[abundance > 0], aes(x = x, y = y, color = abundance)) +
  coord_sf(xlim = c(-2e5, 2.1e5), ylim = c(3.6e6, 4.35e6)) +
  scale_color_viridis_c(trans = 'log10') +
  labs(x = NULL, y = NULL, color = 'Abundance') +
  facet_wrap(~ cut_7y, ncol = 4) +
  theme_minimal()
```

## Variable selection

Unfortunately, not all variables were recorded at every trawl. If we'd like to include variables such as surface and bottom salinity, we'll have to restrict things to after 1992. For now, I'm going to avoid those variables to keep the longest time series possible.

```{r}
obs <- station_key[, lapply(.SD, function(.) sum(!is.na(.))/ length(.)),
                   by = c('year', 'season'), .SDcols = c(11, 13:17)]
obs <- melt(obs, id.vars = c('year', 'season'))

ggplot(data = obs) +
  geom_line(aes(x = year, y = value, color = season)) +
  facet_wrap(~ variable) +
  labs(y = 'Proportion of records available') +
  theme_bw()
```

## GAMM

### Spatial autocorrelation (model mesh)

I'm not an [R-INLA](https://www.r-inla.org/) expert. In fact, I'm pretty well set in the novice category -- most of my modeling experience leverages the [`lme4`](https://github.com/lme4/lme4) package for GLMMs and the [`mgcv`](https://cran.r-project.org/web/packages/mgcv/index.html) package for GAMMs. Luckily, @miller2020 outline a way to cross the boundary between `INLA` and `mgcv` and account for spatial correlation by fitting a stochastic partial differential equation (SPDE) as a smooth. They also do the hard work of coding the smooth, so thank you to them.

Without further ado, here's the mesh:

```{r}
strata[, geometry := st_simplify(geometry,
                                 preserveTopology = F,
                                 dTolerance = 5000)]

library(INLA)

boundary <- inla.sp2segment(as_Spatial(strata$geometry))
# make mesh using meshbuilder 'cause it's finnicky
# meshbuilder()

mesh <- inla.mesh.2d(boundary=boundary,
                     max.edge=c(22000, 108000),
                     min.angle=c(30, 21),
                     
                     ## Safeguard against large meshes.
                     max.n=c(48000, 16000),
                     
                     ## Don't build a huge mesh!
                     max.n.strict=c(128000, 128000),
                     
                     ## Filter away adjacent points.
                     cutoff=8000,
                     offset=c(1000, 55000))

plot(mesh, asp = 1)
```

### Model fitting

To start simple, I'm going to fit a basic model of the following form:

$$
Abundance = \beta_0 + \beta_{season} + \beta_{sediment} + f(bottom temperature) + f(Longitude, Latitude) + f(year)
$$

where $\beta_0$ is the intercept, $\beta_{season}$ is the effect of season, $\beta_{sediment}$ is the effect of sediment type, $f(bottom temperature)$ is a smoothed effect of bottom temperature, $f(Longitude, Latitude)$ is as smoothed effect of space using the mesh above, and $f(year)$ is a smoothed function of year.

I'm going to use all of the trawls available -- including those that did not observe a croaker -- after 1967 (the first year to record a croaker) up to and including 2014. 2015-2019 will be held out for testing. That will use all of these points:

```{r}
croaker <- croaker[complete.cases(croaker[, .(season, bottemp, grpsed,
                                              year, lat, lon)]) &
                     selected == 'Selected' &
                     year >= 1967]

ggplot() +
  geom_sf(data = coast, color = 'lightgray') +
  geom_sf(data = croaker, aes(geometry = geometry), size = 0.1) +
  geom_sf(data = strata, aes(geometry = geometry),
          fill = 'pink', alpha = 0.9) +
  coord_sf(xlim = c(-2e5, 4.5e5), ylim = c(3.6e6, 4.38e6)) +
  theme_minimal()
```

Next, I'm going to play with selection of the error family. I'll look at negative binomial, poisson, Tweedie, and zero-inflated Poisson. Here are the results of AIC:

```{r}
source("miller et al. supp/mgcv_spde_smooth.R") 

library(mgcv)

croaker[, ':='(yr_fac = as.character(year),
               X = st_coordinates(geometry)[, 1],
               Y = st_coordinates(geometry)[, 2])]

abun <- readRDS('data derived/model output/bam_abun_bottemp_xy_yrran_with0_cryr.RDS')
lapply(abun, AIC)
```

Negative binomial is much more supported than the others using AIC.

### Negative binomial summary

```{r}
summary(abun$nb, re.test = F)
```

In the negative binomial model, it seems that the parametric effects of season and sediment group are non-significant. There is high deviance explained here, but that is almost definitely due to the number of zeroes in the data. Obviously, the R^2^ value here is nonsense, and likely suggests that we're missing some process in the model. When running a model on presence-only abundance, deviance explained is \~ 35% and R^2^ is 4.41%.

### Negative binomial visualization

```{r}
preds <- predict(abun$nb, type = 'terms', se = T)

preds <- setDT(cbind(
  abun$nb$model,
  pred = exp(preds$fit),
  lci = exp(preds$fit - 1.96 * preds$se.fit),
  uci = exp(preds$fit + 1.96 * preds$se.fit)
))


ggplot(data = preds, aes(x = bottemp)) +
  geom_ribbon(aes(ymin = `lci.s(bottemp)`, ymax = `uci.s(bottemp)`),
              fill = 'lightgray') +
  geom_line(aes(y = `pred.s(bottemp)`)) +
  geom_rug() +
  labs(x = 'Bottom temperature (°C)', y = 'Individuals',
       title = 'Partial effect of bottom temperature') +
  theme_bw()

```

```{r}
ggplot(data = preds, aes(x = year)) +
  geom_ribbon(aes(ymin = `lci.s(year)`, ymax = `uci.s(year)`),
              fill = 'lightgray') +
  geom_line(aes(y = `pred.s(year)`)) +
  labs(x = NULL, y = 'Individuals',
       title = 'Partial effect of year') +
  theme_bw()
```

```{r}
grid <- st_make_grid(strata$geometry, cellsize = 10*1000)

grid <- grid[st_covered_by(grid, strata$geometry, sparse = F) |
                st_overlaps(grid, strata$geometry, sparse = F)]

grid_pts <- st_centroid(grid)
grid_pts <- data.table(X = st_coordinates(grid_pts)[,1],
                       Y = st_coordinates(grid_pts)[,2],
                       season = c('FALL', 'SPRING'),
                       grpsed = c('Sand', 'Gravel'),
                       bottemp = 20, year = 1973)

spat_pred <- predict(abun$nb, grid_pts,
                     type = 'terms', terms = 's(X,Y)', se = T,
                     newdata.guaranteed = T)
spat_pred <- data.table(
  data.table(grid),
  pred = exp(spat_pred$fit),
  lci = exp(spat_pred$fit - 1.96 * spat_pred$se.fit),
  uci = exp(spat_pred$fit + 1.96 * spat_pred$se.fit)
)


spat_pred <- melt(spat_pred,
                 id.vars = 'grid',
                 measure.vars = c('pred.s(X,Y)', 'lci.s(X,Y)', 'uci.s(X,Y)'))
spat_pred[, variable := fcase(variable %like% 'pred', 'Predicted',
                             variable %like% 'lci', 'Lower confidence interval',
                             variable %like% 'uci', 'Upper confidence interval')]

ggplot() +
  geom_sf(data = spat_pred, aes(geometry = grid, fill = value), color = NA) +
  geom_sf(data = coast) +
  coord_sf(xlim = c(-2e5, 4.2e5), ylim = c(3.6e6, 4.5e6)) +
  scale_fill_viridis_c(trans = 'log10') +
  labs(x = NULL, y = NULL, fill = NULL,
       title = 'Atlantic croaker modeled abundance',
       subtitle = 'Partial effect of X-Y space') +
  facet_wrap(~variable) +
  theme_minimal() +
  theme(legend.position = c(0.9, 0.4))
```

### Cross validation

Since this takes a lot of computing time, cross validation was constructed "off screen". I ran 5-fold cross validation in two ways:

1)  The "canonical" way for time series, where the model was

    1.  fit with data up to 2014, then tested on 2015 observations
    2.  fit with data up to 2015, then tested on 2016 observations
    3.  fit with data up to 2016, then tested on 2017 observations
    4.  and so on through 2018. This tests the model's predictive ability in the next year.

2)  Fit the model with data up to 2014, then test this model on observations each year through 2019.

    1.  This tests the model's predictive ability through five years in the future.

#### Canonical cross validation

The mean root-mean-square error of predicting the next year's abundance when catch was greater than zero was 815 individuals,. This was about 25% of each year's maximum value. When catch was 0, the model it it more-or-less bang on. This tells us that the zeroes are really in the driver's seat here.

```{r}
cv <- fread('data derived/model output/canonical_cv_croaker_abun.csv')

cv[, .('RMSE of folds' = round(mean(rmse), 2),
       'Pct. max' = round(median(prop_max) * 100, 0)),
   by = 'abundance > 0']
```

#### 5 year progressive cross validation

Running 5-fold cross-validation in the second manner yields nearly identical results to the first. It seems that that the parameters within the model have stabilized using data up until 2014; additional data does not change the predictive error of the model.

```{r}
cv <- fread('data derived/model output/5yr_cv_croaker_abun.csv')

cv[, .('RMSE of folds' = round(mean(rmse), 2),
       'Pct. max' = round(median(prop_max) * 100, 0)),
   by = 'abundance > 0']
```

## References
