---
params:
  species: !r
title: "`r paste(params$species, 'distribution through time <br>50 km x 50 km grid')`"
output: ioslides_presentation

---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, error = TRUE)
```


```{r}
plot_fun <- function(yr){
  sub_plot <- function(yr, ssn){
    ggplot() +
      
      geom_sf(data = com_spec[year == yr & season == ssn],
              aes(geometry = polygons)) +
      geom_sf(data = abun_grid[year == yr & season == ssn],
              aes(geometry = geometry, fill = med_abun), color = NA) +
      geom_sf(data = com_spec[year == yr & season == ssn], 
              aes(geometry = geometry), size = 5, shape = 8) +
      geom_sf(data = com_spec[year == yr & season == ssn],
              aes(geometry = com_dist_line), color = 'red') +
      geom_sf(data = coast) +
      coord_sf(xlim = c(-4e5, 6.7e5), ylim = c(3.4e6, 4.6e6)) +
      scale_fill_viridis_c(option = 'plasma') +
      labs(subtitle = ssn, fill = 'Med. \nabund.') +
      theme_minimal() +
      theme(legend.position = c(0.9, 0.25))
  }
  
  sub_plot(yr, 'SPRING') + sub_plot(yr, 'FALL')

}

```



## Longitude center of mass (weighted by abundance) through time 
```{r}
ggplot(data = com_spec, aes(x = year, y = x_wt/1000)) +
  geom_line() +
  coord_flip() + 
  geom_smooth(method = 'lm') +
  facet_wrap(~ season) +
  labs(x = NULL, y = 'Eastings (km)') +
  theme_bw()
```


## Latitude center of mass (weighted by abundance) through time
```{r}
ggplot(data = com_spec, aes(x = year, y = y_wt/1000)) +
  geom_line() +
  geom_smooth(method = 'lm') +
  facet_wrap(~season) +
  labs(x = NULL, y = 'Northings (km)') +
  theme_bw()
```


## Distance from shore of the center of mass through time
```{r}
ggplot(data = com_spec, aes(x = year, y = com_dist / 1000)) +
  geom_line() +
  geom_smooth(method = 'lm') +
  labs(x = NULL, y = 'Distance from shore (km)') +
  facet_wrap(~season) +
  theme_bw()
```


## Distance-from-shore center of mass (site distance weighted by abundance) through time
```{r}
ggplot(data = com_spec, aes(x = year, y = dist_wt / 1000)) +
  geom_line() +
  geom_smooth(method = 'lm') +
  labs(x = NULL, y = 'Distance from shore (km)') +
  facet_wrap(~season) +
  theme_bw()
```

## Depth center of mass (weighted by abundance) through time
```{r}
ggplot(data = com_spec, aes(x = year, y = dep_wt)) +
  geom_line() +
  geom_smooth(method = 'lm') +
  labs(x = NULL, y = 'Depth (ft)') +
  facet_wrap(~season) +
  theme_bw()
```


## Bottom temperature center of mass (weighted by abundance) through time
```{r}
ggplot(data = com_spec, aes(x = year, y = bt_wt)) +
  geom_line() +
  geom_smooth(method = 'lm') +
  labs(x = NULL, y = 'Bottom temperature (C)') +
  facet_wrap(~season) +
  theme_bw()
```


## Area of concave hull through time
```{r}
ggplot(data = com_spec, aes(x = year, y = as.numeric(st_area(polygons))/1e6)) +
  geom_line() +
  geom_smooth(method = 'lm') +
  labs(x = NULL, y = 'Area (km^2)') +
  facet_wrap(~season) +
  theme_bw()
```


```{r, results="asis"}
for(i in seq(1967, 2019, 1)){
  
  cat("\n") 
  cat("##", i, "\n") # Create new slides with the names.
  
  print(
    
    plot_fun(i)
    
  )
  
  cat("\n")
  
}

```